---
title: "An Exploratory Look Aboard the Titanic"
author: "Marshall McQuillen"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE)
```

## Guiding Question

Which passengers will survive the sinking of the Titanic?

## Secondary Questions

1. What characteristics separate those who survived from those who died?
2. What charactersistics make someone more likely to survive?
3. How do different characteristics of passengers vary with one another?

```{r import_data}
setwd("/Users/marsh/data_science_projects/Kaggle_Competitions/titanic_survival_classification/")
training <- read.csv('/Users/marsh/data_science_projects/Kaggle_Competitions/titanic_survival_classification/titanic_train.csv')
```

```{r install/load_packages}
# install.packages("tidyverse")
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(knitr))
```

## Data Overview

Looking at the training data froma bird-eye view, there are 891 observations representing passengers and 12 variables. Since some of the variable name are a little cryptic, an description for each is provided below.

| Variable Name | Description |
| ----------------------------------------- | ----------------------------------------- |
| PassengerId | Unique identifier for each passenger |
| Survived | Binary; 1 = Survied & 0 = Died |
| Pclass | Socio-economic status; 1 = Upper, 2 = Middle & 3 = Lower |
| Name | Passenger Name |
| Sex | Male or Female |
| Age | Passenger Age |
| SibSp | Number of siblings or spouse aboard ship |
| Parch | Number of parents or children aboard ship |
| Ticket | Ticket Number |
| Fare | Amount paid for ticket |
| Cabin | Cabin number |
| Embarked | The town from which the passenger boarded the ship; C = Cherbourg, Q = Queenstown & S = Southhampton |

First and foremost, by running `str(training)` on the data, it is apparent that the first entries in the Cabin and Embarked columns are empty strings, indicating that the data is probably not perfectly clean (no surprises there). Checking to see where any Null's might be, it becomes clear that there are in fact no nulls, and that these spaces were intentionally left empty. In addition to null values, all the NA's are in the Age, accounting for roughly 20% of the values in that column. Both of these will need to be imputed intelligently when the time to create a predictive model comes around.

In addition to the missing values, it is important to note that some of the discrete attributes have been read in as continous variabes such as Pclass, Sibsp and Parch. Since these variables actually represent discrete characteristics of each passenger, changing them to be non-continuous will allow a more representative analysis.

```{r data_overview}
# str(training)

#Set classes
training$PassengerId <- as.factor(training$PassengerId)
training$Survived <- as.factor(training$Survived)
training$Pclass <- as.factor(training$Pclass)
training$SibSp <- as.factor(training$SibSp)
training$Parch <- as.factor(training$Parch)

# view null count
nulls <- lapply(training[,1:12], is.null)
na <- lapply(training[,1:12], is.na)
null_count <- lapply(nulls, sum)
na_count <- lapply(na, sum)

df <- mapply(data.frame, null_count, na_count)
row.names(df) <- c("Null Count", "NA Count")

df1 <- df[,1:6]
df2 <- df[,7:12]

knitr::kable(df1, caption = "Attribute Null & NA Counts")
knitr::kable(df2, caption = "Attribute Null & NA Counts (continued)")

#Percent of Age variable that is NA
age_percent_na <- paste(round(sum(is.na(training$Age))/length(training$Age)*100, digits = 2), "%", sep = "")
```

## 2 Bayesian Survival

### 2.2 Does Money Sink or Swim?

By creating a table with the Pclass (which, recall refers to the socioeconomic status (SES) of the passenger) and Survived variables, I can get a good sense of the number of passengers that lived and died, based on their SES. Simple summation and division returns the probabilites of a passenger living given their respective SES

```{r pclass_vs_survived}
#Probability of living by socio-economic status
pclass_table <- with(training, table(Survived, Pclass))
upper_class <- pclass_table[2,1]/sum(pclass_table[,1])*100
middle_class <- pclass_table[2,2]/sum(pclass_table[,2])*100
lower_class <- pclass_table[2,3]/sum(pclass_table[,3])*100

row.names(pclass_table) <- c('Died','Survived')
colnames(pclass_table) <- c('Upper','Middle','Lower')
knitr::kable(pclass_table, caption = "Survival Counts by SES")
#Probability of living given Upper Class
upper_sr <- paste(round(upper_class, digits = 2), "%", sep = "")

#Probability of living given Middle Class
middle_sr <- paste(round(middle_class, digits = 2), "%", sep = "")

#Probability of living given Lower Class
lower_sr <- paste(round(lower_class, digits = 2), "%", sep = "")

sr_table <- data.frame(c(upper_sr, middle_sr, lower_sr), row.names = c("Upper Class",'Middle Class','Lower Class'))
colnames(sr_table) <- 'Probability of Surviving'
knitr::kable(sr_table, caption = "Survival Rates by SES", align = 'r')
```

This same information is displayed visually below.

```{r bayes_part_2}
g <- ggplot(training, aes(y=Survived,  x=factor(Survived,
                                                labels=c("Died","Lived"))))
g <- g + geom_bar(aes(y=..prop.., group=Pclass,
                      fill=factor(..x.., labels=c("Died","Lived"))))
g <- g + facet_grid(~factor(Pclass,
                            labels=c("Upper Class", "Middle Class",
                                     "Lower Class")))
g <- g + scale_y_continuous(labels = scales::percent) 
g <- g + scale_fill_discrete(name="Survival Status")
g <- g + labs(x="", y = "Percentage",
              title = "Probabilities of Living Given Socio-Economic Status")
g <- g + geom_text(
      aes(label = paste(round((..count../c(216,216,184,184,491,491)), 4)*100, "%", sep = ""), y = ..prop..),  stat= "count",
      vjust = c(17.25,10,12.75,14.5,6.5,21))
g
```

#### Illustrating Bayes Theorem with Survival Rates and Socio-Economic Status

This type of classification problem creates a great opportunity to illustrate Bayes' Theorem. Recall that Bayes Theorem is defined as:

$$P(A|B)~=~\frac{P(B|A)P(A)}{P(B)}$$



...I can set **P(A) = the probability of a passenger belonging to a defined SES (X)** and **P(B) = the probability of a passenger living**. I can now rewrite the previously defined Theorm using my definitions as:  

$$P(~"X~class~citizen"~|~"Lived"~)~=~\frac{P(~"Lived"~|~"X~class~citizen"~)~P(~"X~class~citizen"~)}{(~"Lived"~)}$$

The same information can be displayed visually as follows.





```{r bayes_part_3}
total <- nrow(training)
total_died <- nrow(subset(training, Survived == 0))
total_lived <- nrow(subset(training, Survived == 1))

#Probability of Living = P(B)
prob_lived <- total_lived/(total_died + total_lived)

#Probability of being Upper, Middle or Lower class = P(A)
upper_prob <- as.numeric(table(training$Pclass)[1])/total
middle_prob <- as.numeric(table(training$Pclass)[2])/total
lower_prob <- as.numeric(table(training$Pclass)[3])/total
```

Now that I have found both **P("X class citizen")** (objects upper_prob, middle_prob and lower_prob) and **P("Lived")** (object prob_lived), and I have **P("Lived"|"X class ctiizen")** (objects upper_class, middle_class and lower_class), I can solve for **P("X class citizen"|"Lived")**...

<br>
<center> <h5> P( "Upper class citizen" | "Lived" ) = <u>upper_class X upper_prob</u> <br><p style="margin-left: 250px">prop_lived</p></br> </h5> </center>

<center> <h5> P( "Middle class citizen" | "Lived" ) = <u>middle_class X middle_prob</u> <br><p style="margin-left: 250px">prop_lived</p></br> </h5> </center>

<center> <h5> P( "Lower class citizen" | "Lived" ) = <u>lower_class X lower_prob</u> <br><p style="margin-left: 250px">prop_lived</p></br> </h5> </center>
</br>

```{r bayes_part_4}
#Probability of being an Upper class citizen given a passenger lived
prob_upper_given_lived <- (upper_class*upper_prob)/prob_lived

#Probability of being a Middle class citizen given a passenger lived
prob_middle_given_lived <- (middle_class*middle_prob)/prob_lived

#Probability of being a Lower class ctizen given a passenger lived
prob_lower_given_lived <- (lower_class*lower_prob)/prob_lived
```

The "shorthand" way of finding these probabilites can be accomplished by dividing the the number of X class passengers that lived by the total number of passengers that lived using the pclass_table.

```{r}
#"Shorthand" for calculating probability of being a X class citizen using pclass_table
upper_class_by_total <- pclass_table[2,1]/total_lived*100
middle_class_by_total <- pclass_table[2,2]/total_lived*100
lower_class_by_total <- pclass_table[2,3]/total_lived*100

#Showing that the probabilities using Bayes Theorem and pclass_table are equal
all.equal(upper_class_by_total, prob_upper_given_lived)
all.equal(middle_class_by_total, prob_middle_given_lived)
all.equal(lower_class_by_total, prob_lower_given_lived)

```

A graphic illustrating said results.

```{r}
f <- ggplot(training, aes(y = Pclass,
                          x = factor(Pclass,labels=c("Upper","Middle","Lower"))))
f <- f + geom_bar(aes(y=..prop.., group=Survived,
                      fill = factor(..x.., labels=c("Upper Class","Middle Class","Lower Class"))))
f <- f + facet_grid(~factor(Survived, labels = c("Died","Lived")))
f <- f + scale_y_continuous(labels = scales::percent)
f <- f + scale_fill_discrete(name = "Socio-Economic Status")
f <- f + labs(x = "", y = "Percentage", title = "Probabilities of Socio-Economic Status Given Died or Survived")
f <- f + geom_text(
      aes(label = paste(round((..count../c(549,549,549,342,342,342)), 4)*100, "%", sep = ""), y = ..prop..), stat= "count",
      vjust = c(24,23,9,16.75,20.75,18.25))
f

```

Using the same code as above, with a few minor adjustments I can make similar graphs with other qualitative variables, such as the sex of the passenger.

```{r sex/survived}
#Sex vs. Survived
a <- ggplot(training, aes(y=Survived,  x=factor(Survived,
                                                labels=c("Died","Lived"))))
a <- a + geom_bar(aes(y=..prop.., group=Sex,
                      fill=factor(..x.., labels=c("Died","Lived"))))
a <- a + facet_grid(~factor(Sex,
                            labels=c("Female", "Male")))
a <- a + scale_y_continuous(labels = scales::percent) 
a <- a + scale_fill_discrete(name="Survival Status")
a <- a + labs(x="", y = "Percentage",
              title = "Probabilities of Living Given Gender")
a <- a + geom_text(
      aes(label = paste(round((..count../c(314,314,577,577)), 4)*100, "%", sep = ""), y = ..prop..),  stat= "count",
      vjust = c(20.5,7,5,22.5))
a
```

## 3 Cabin Classification

It seems logical that looking at *where* each passenger was when the Titanic started sinking could provide some insight as to why some lived and others did not. The "Sinking" section on the [Titanic Wikipedia Page](https://en.wikipedia.org/wiki/RMS_Titanic) states that the iceberg was struck at 11:40 pm. Considering the time of night, combined with the likely cold air temperature, I think it is safe to say that most passengers were inside, if not in their rooms sleeping.

Finding out where each passenger was will be a two fold process:

1. Subsetting on the Deck they were on, noted by the letter in the Cabin column.
2. Subsetting where on that deck they were, noted by the number in the Cabin column.

An important note is that the vast majority of the passengers did not have an entry in the Cabin column. (There aren't any NA's, the entries are not even filled with spaces, they are simply "nothing"). In order to subset these observations, I used the ouput from a "nothing" observation in the logical statement.

After subsetting, summing the number of rows in each subset, *which should equal 891, the total number of observations,* returns 894. A little searching led to finding the duplicates, show below.


```{r}
#Split data on Cabin Letter
a_class <- training[grep("A", training$Cabin),]
b_class <- training[grep("B", training$Cabin),]
c_class <- training[grep("C", training$Cabin),]
d_class <- training[grep("D", training$Cabin),]
e_class <- training[grep("E", training$Cabin),]
f_class <- training[grep("F", training$Cabin),]
g_class <- training[grep("G", training$Cabin),]

#the "nothing" class
blank_class <- subset(training, Cabin == training[1,11])

sum(nrow(a_class) + nrow(b_class) + nrow(c_class) + nrow(d_class) +
          nrow(e_class) + nrow(f_class) + nrow(g_class) + nrow(blank_class))

#Duplicate Cabin Values
duplicates <- training[c(76,129,700,716),]
duplicates

#look at passenger id 534
training[grep("Peter", training$Name),]
```

To decide which subset to assign these observations too, looking at the Embarked and Ticket columns for those observations in the g_class subset, I can see that everyone in this cabin class embarked from Southampton and had similar ticket


```{r}
table(a_class$Survived)[2]/sum(table(a_class$Survived))
table(b_class$Survived)[2]/sum(table(b_class$Survived))
table(c_class$Survived)[2]/sum(table(c_class$Survived))
table(d_class$Survived)[2]/sum(table(d_class$Survived))
table(e_class$Survived)[2]/sum(table(e_class$Survived))
table(f_class$Survived)[2]/sum(table(f_class$Survived))
table(g_class$Survived)[2]/sum(table(g_class$Survived))
table(blank_class$Survived)[2]/sum(table(blank_class$Survived))

#Split data on Cabin Room Number
tCabin_number = grep("[0-9]{2,}", training$Cabin)
test <- mutate(training, test_column = c(strsplit(as.character(training$Cabin), " ")))

#Parch and Sibsp Analysis
table(training$Parch, training$Survived)

h <- ggplot(data = training,
            aes(y = Survived,
                x = factor(Survived, labels = c("Died","Lived"))))
h <- h + geom_bar(aes(y = ..prop.., group = Parch,
                      fill = factor(..x.., labels = c("Died","Lived"))))
h <- h + facet_grid(~Parch)
h <- h + scale_y_continuous(labels = scales::percent)
h <- h + scale_fill_discrete()
h

d <- ggplot(data = training,
            aes(y = Survived,
                x = factor(Survived, labels = c("Died","Lived"))))
d <- d + geom_bar(aes(y = ..prop.., group = SibSp,
                      fill = factor(..x.., labels = c("Died","Lived"))))
d <- d + facet_grid(~SibSp)
d <- d + scale_y_continuous(labels = scales::percent)
d <- d + scale_fill_discrete()
d

```



